---
title: "DS311 - R Lab Assignment"
author: "Jessica Torres"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: united
    highlight: tango
    df_print: paged
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Assignment 1

* In this assignment, we are going to apply some of the build in data set in R for descriptive statistics analysis.
* To earn full grade in this assignment, students need to complete the coding tasks for each question to get the result.
* After finished all the questions, knit the document into HTML format for submission.

### Question 1

Using the **mtcars** data set in R, please answer the following questions.

```{r}
# Loading the data
data(mtcars)

# Head of the data set
head(mtcars)
```

a. Report the number of variables and observations in the data set.

```{r}
# Enter your code here!
dim(mtcars)

# Answer:
print("There are total of _____ variables and _____ observations in this data set.")
```



b. Print the summary statistics of the data set and report how many discrete and continuous variables are in the data set.

```{r}
# Enter your code here!
summary(mtcars)

# Answer:
print("There are _____ discrete variables and _____ continuous variables in this data set.")
```



c. Calculate the mean, variance, and standard deviation for the variable **mpg** and assign them into variable names m, v, and s. Report the results in the print statement.

```{r}
# Enter your code here!
m <- colMeans(mtcars["mpg"])
v <- var(mtcars["mpg"])
s <- sd(as.numeric(unlist(mtcars["mpg"])))



# print(paste("The average of Mile Per Gallon from this data set is", m , " with variance ", v , " and standard deviation", s , "."))
```



d. Create two tables to summarize 1) average mpg for each cylinder class and 2) the standard deviation of mpg for each gear class.

```{r}
# Enter your code here!
# Load the dplyr package if not already loaded
if (!require(dplyr)) {
  install.packages("dplyr")
  library(dplyr)
}

# 1) Average mpg for each cylinder class
avg_mpg_by_cyl <- mtcars %>%
  group_by(cyl) %>%
  summarise(avg_mpg = mean(mpg))

# Print the result
print("Table 1: Average mpg for each cylinder class")
print(avg_mpg_by_cyl)

# 2) Standard deviation of mpg for each gear class
sd_mpg_by_gear <- mtcars %>%
  group_by(gear) %>%
  summarise(sd_mpg = sd(mpg))

# Print the result
print("Table 2: Standard deviation of mpg for each gear class")
print(sd_mpg_by_gear)





```



e. Create a crosstab that shows the number of observations belong to each cylinder and gear class combinations. The table should show how many observations given the car has 4 cylinders with 3 gears, 4 cylinders with 4 gears, etc. Report which combination is recorded in this data set and how many observations for this type of car.

```{r}
# Enter your code here!
# Create a cross-tabulation of cylinder and gear classes
cross_tab <- table(mtcars$cyl, mtcars$gear)

# Print the result
print("Cross-tabulation of Cylinder and Gear Classes:")
print(cross_tab)
# Extract row and column names
row_names <- rownames(cross_tab)
col_names <- colnames(cross_tab)

# Report the combinations and counts
for (i in 1:length(row_names)) {
  for (j in 1:length(col_names)) {
    count <- cross_tab[i, j]
    print(paste("Cylinder:", row_names[i], "Gears:", col_names[j], "Count:", count))
  }
}


print("The most common car type in this data set is car with ____ cylinders and ____ gears. There are total of _____ cars belong to this specification in the data set.")
```


*** 

### Question 2

Use different visualization tools to summarize the data sets in this question.

a. Using the **PlantGrowth** data set, visualize and compare the weight of the plant in the three separated group. Give labels to the title, x-axis, and y-axis on the graph. Write a paragraph to summarize your findings.

```{r}
# Load the data set
data("PlantGrowth")

# Head of the data set
head(PlantGrowth)

# Enter your code here!
boxplot(weight ~ group, data = PlantGrowth, 
        main = "Plant Weight by Group",
        xlab = "Group",
        ylab = "Weight")

# Add labels for each box
box_labels <- levels(PlantGrowth$group)
text(seq_along(box_labels), boxplot(weight ~ group, data = PlantGrowth, plot = FALSE)$stats[3, ] + 1, box_labels, pos = 3, col = "blue")

```

Result:

=> Report a paragraph to summarize your findings from the plot!



b. Using the **mtcars** data set, plot the histogram for the column **mpg** with 10 breaks. Give labels to the title, x-axis, and y-axis on the graph. Report the most observed mpg class from the data set.

```{r}

# Plot histogram for mpg with 10 breaks
hist(mtcars$mpg, breaks = 10, main = "Histogram of MPG in mtcars", xlab = "MPG", ylab = "Frequency", col = "skyblue")

# Identify the most observed MPG class
most_observed_mpg <- cut(mtcars$mpg, breaks = 10)
most_observed_mpg_class <- levels(most_observed_mpg)[which.max(table(most_observed_mpg))]
print(paste("The most observed MPG class is:", most_observed_mpg_class))


print("Most of the cars in this data set are in the class of ______ mile per gallon.")
```



c. Using the **USArrests** data set, create a pairs plot to display the correlations between the variables in the data set. Plot the scatter plot with **Murder** and **Assault**. Give labels to the title, x-axis, and y-axis on the graph. Write a paragraph to summarize your results from both plots.

```{r}
# Load the data set
data("USArrests")

# Head of the data set
head(USArrests)

# Enter your code here!
pairs(USArrests, main = "Pairs Plot for USArrests Dataset")

# Scatter plot for Murder and Assault
plot(USArrests$Murder, USArrests$Assault, 
     main = "Scatter Plot: Murder vs. Assault",
     xlab = "Murder", ylab = "Assault", col = "blue", pch = 16)

```

Result:

=> Report a paragraph to summarize your findings from the plot!



***

### Question 3

Download the housing data set from www.jaredlander.com and find out what explains the housing prices in New York City.

Note: Check your working directory to make sure that you can download the data into the data folder. 

```{r, echo=FALSE}
# Load and clean the housing data set 
download.file(url='https://www.jaredlander.com/data/housing.csv',
              destfile='data/housing.csv', mode='wb')
housingData <- read.csv('data/housing.csv')
housingData <- subset(housingData, 
                      select = c("Neighborhood", "Market.Value.per.SqFt", "Boro", "Year.Built"))
housingData <- na.omit(housingData)                   
```

a. Create your own descriptive statistics and aggregation tables to summarize the data set and find any meaningful results between different variables in the data set.

```{r}
# Head of the cleaned data set
head(housingData)

# Enter your code here!
# Load necessary libraries
library(dplyr)

# Descriptive statistics for numeric variables
numeric_summary <- housingData %>%
  select(Market.Value.per.SqFt, Year.Built) %>%
  summarise(
    Mean_Market_Value = mean(Market.Value.per.SqFt),
    Median_Year_Built = median(Year.Built),
    Min_Market_Value = min(Market.Value.per.SqFt),
    Max_Year_Built = max(Year.Built)
  )

# Aggregation table by Boro
boro_aggregation <- housingData %>%
  group_by(Boro) %>%
  summarise(
    Mean_Market_Value = mean(Market.Value.per.SqFt),
    Median_Year_Built = median(Year.Built),
    Num_Observations = n()
  )

# Aggregation table by Neighborhood
neighborhood_aggregation <- housingData %>%
  group_by(Neighborhood) %>%
  summarise(
    Mean_Market_Value = mean(Market.Value.per.SqFt),
    Median_Year_Built = median(Year.Built),
    Num_Observations = n()
  )

# Print the results
print("Descriptive Statistics:")
print(numeric_summary)

print("Aggregation by Boro:")
print(boro_aggregation)

print("Aggregation by Neighborhood:")
print(neighborhood_aggregation)


```

b. Create multiple plots to demonstrates the correlations between different variables. Remember to label all axes and give title to each graph.

```{r}
# Enter your code here!
install.packages("ggplot2")
install.packages("corrplot")

# Load necessary libraries
library(ggplot2)
library(corrplot)

# Scatter plot: Market Value per SqFt vs. Year Built
ggplot(housingData, aes(x = Year.Built, y = Market.Value.per.SqFt)) +
  geom_point() +
  labs(title = "Scatter Plot: Market Value per SqFt vs. Year Built",
       x = "Year Built",
       y = "Market Value per SqFt")

# Scatter plot matrix
scatter_matrix <- ggplot(housingData, aes(x = Year.Built, y = Market.Value.per.SqFt)) +
  geom_point() +
  facet_grid(. ~ Boro) +
  labs(title = "Scatter Plot Matrix: Market Value per SqFt vs. Year Built",
       x = "Year Built",
       y = "Market Value per SqFt")

# Correlation matrix
cor_matrix <- cor(housingData[, c("Market.Value.per.SqFt", "Year.Built")])
corrplot(cor_matrix, method = "circle", title = "Correlation Matrix")

# Print the plots
print(scatter_matrix)


```

c. Write a summary about your findings from this exercise.

=> Enter your answer here!
In this analysis of a housing dataset from New York City, I conducted exploratory data analysis to understand the factors influencing market value per square foot. Descriptive statistics provided insights into the central tendency and variability of the dataset, while aggregation tables by boroughs and neighborhoods allowed for geographic comparisons. Scatter plots and correlation matrices visually represented relationships between variables, revealing a positive correlation between market value and the year a building was built. These preliminary findings set the stage for more advanced analyses, offering a foundation for understanding the complex dynamics driving housing prices in the city.